---
title: "Data Wrangling: Intro to `tidyr`"
author: "Emorie D Beck"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    code_folding: show
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = F)
```

```{r}
library(psych)
library(tidyverse)
```

# Review  

In the last tutorial, we covered a series of functions from the `dplyr` package that are useful for data manipulation, including `filter()`, `select()`, `arrange()`, `group_by()`, `mutate()`, and `summarize()` / `summarise()`.  

# `tidyr`
Today, we'll build off these useful functions for data manipulation with useful tools from the `tidyr` that are fantastic for data *wrangling*. Specifically, we'll cover:  

1. `pivot_longer()`, which takes a "wide" format data frame and makes it long.  
2. `pivot_wider()`, which takes a "long" format data frame and makes it wide.  

Then, we'll actually step back and cover a few additional functions from the `dplyr` package that are useful for merging, or joining, different data frames together, including:  

3. `full_join()`, which merges *all* rows in either data frame provided as arguments.  
4. `inner_join()`, which merges rows whose keys are present in *both* data frames provided as arguments.  
5. `left_join()`, which "prioritizes" the first data set provided as an argument.  
6. `right_join()`, which "prioritizes" the second data set provided as an argument.  

There are a few additional join functions within `dplyr`, including`anti_join()` and `semi_join()`, which have utility as well, but I'll just direct you to their documentation to keep this from getting overwhelming.  

# Key `tidyr` Functions  
## 1. `pivot_longer()`  

## 2. `pivot_wider()`  

# More `dplyr` Functions  
## 3. `full_join()`  

## 4. `inner_join()`  

## 5. `left_join()`  

## 6. `right_join()`  


