---
title: "`purrr`"
author: "Emorie D Beck"
date: "February 21, 2019"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    code_folding: show
    toc: true
    toc_float: true
---

```{r, echo = F}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, error = F,
                     class.source = "code-style")
```

# Iteration  
Iteration is everywhere. It underpins much of mathematics and statistics. If you've ever seen the $\Sigma$ symbol, then you've seen (and probably used) iteration.  

Reasons for iteration:  
- reading in multiple files from a directory  
- running the same operation multiple times  
- running different combinations of the same model  
- creating similar figures / tables / outputs  

```{r packages, echo=F}
library(psych)
library(knitr)
library(kableExtra)
library(lme4)
library(broom.mixed)
library(plyr)
library(tidyverse)
```

## `for` loops  
Enter `for` loops. `for` loops are the "OG" form of iteration in computer science. The basic syntax is below. Basically, we can use a for loop to loop through and print a series of things.

```{r basic loop}
for(i in letters[1:5]){
  print(i)
}
```

The code above "loops" through 5 times, printing the iteration letter.  

## `_apply()` family  
A somewhat faster version of `for` loops comes from the `_apply()` family of functions, including: `apply()`, `lapply()`, `sapply()`, and `mapply()`. Unlike `for` loops, these are vectorized, which makes them more efficient.  

```{r apply functions}
lapply(letters[1:5], print)
sapply(letters[1:5], print)
mapply(print, letters[1:5])
```

## `purrr` and `_map_()` functions  
Today, though, we'll focus on the `map()` family of functions, which is the functions through which `purrr` iterates.  

```{r map functions}
map(letters[1:5], print)
```

**For a more thorough comparison of `for` loops, the `_apply()` family, and `_map_()` functions, see https://jennybc.github.io/purrr-tutorial/**

## `purrr` and `_map_()` predicates  
Today, though, we'll focus on the `map()` family of functions, which is the functions through which `purrr` iterates.  

```{r map functions 2, eval = F}
map(letters[1:5], print)
```

Note that this returns a list, which we may not always want. With `purrr`, we can change the kind of output of `map()` by adding a predicate, like `lgl`, `dbl`, `chr`, and `df`. So in the example above, we may have wanted just the characters to print. To do that we'd call `map_chr()`:  

```{r basic map_chr}
map_chr(letters[1:5], print)
```

Note that it also returns the concatenated character vector as well as printing each letter individually (i.e. iteratively).  

## `purrr` and `_map_()` antecedents  

- Single mapping: `map_()`  
- Parallel (2) mapping(s): `map2_()`  
- 3 or more mappings: `pmap_()`  

```{r}
map2_chr(letters[1:5], 1:5, paste)
```

Note here that we can use `map2()` and `pmap()` with the predicates from above.  

# Use Cases  

1. Reading Data  
2. Running Models  
3. (Plotting Figures - Kendra Smith, March 20)   
4. (Creating Tables - Emorie Beck (Me), April 17)  

## Reading Data  
There are a number of different cases where `purrr` and `map()` maybe useful for reading in data including: 

- subject-level files for an experimental task  
- subject- and task-level files for an experimental task
- EMA data  
- longitudinal data  
- web scraping and text mining  

## Reading Data: Subject-Level EMA  
For this first example, I'll show you how this would look with a `for` loop before I show you how it looks with `purrr`.

Assuming you have all the data in a single folder and the format is reasonably similar, you have the following basic syntax:  


```{r simple reading loop, eval = F}
data_path <- ""
files <- list.files(data_path)
data <- list()
for(i in files){
  data[[i]] <- read.csv(i, stringsAsFactors = F)
}
data <- combine(data)
```

This works fine in this simple case, but where `purrr` really shines in when you need to make modifications to your data before combining, whether this be recoding, removing missing cases, or renaming variables. 

But first, the simple case of reading data.  

```{r read data ex1, eval = T}
data_path <- "~/Desktop/wustl_r_workshops/purrr"
df1 <- tibble(
  ID = list.files(sprintf("%s/data/example_1", data_path))
  ) %>%
  mutate(data = map(ID, ~read_csv(sprintf("%s/data/example_1/%s", data_path, .)))) %>%
  unnest(data) 
```

The code above creates a list of ID's from the data path (files named for each person), reads the data in using the `map()` function from `purrr`, removes the ".csv" from the ID variable, then unnests the data, resulting in a data frame for each person.  

# List Columns and the Power of `purrr`  
On the previous slide, we saw a data frame **inside** of a data frame. This is called a list column within a nested data frame.  

In this case, we created a list column using map, but one of the best things about `purrr` is how it combines with the `nest()` and `unnest()` functions from the `tidyr` package.  

We'll return to `nest()` later to demonstrate how anything you would iterate across is also something we can `nest()` by in long format data frames.  

# Exercise: Reading in Subject-Level Data  

```{r codebook}
data_path <- "~/Desktop/wustl_r_workshops/purrr"
codebook <- sprintf("%s/data/codebook_ex1.csv", data_path) %>% read_csv(.)
codebook
```

Now, we're going to combine with what we learned about last time with codebooks.  

Now, that we have a codebook, what are the next steps?  
1. pull old names in raw data from codebook  
2. pull new names from codebook  
3. select columns from codebook in loaded data  
4. rename columns with new names  

# Exercise: Reading in Subject-Level Data  
```{r read data complex ex1, eval = T}
old.names <- codebook$old_name # pull old names in raw data from codebook  
new.names <- codebook$new_name # pull new names from codebook  
df1 <- tibble(
  ID = list.files(sprintf("%s/data/example_1", data_path))
  ) %>%
  mutate(data = map(ID, ~read_csv(sprintf("%s/data/example_1/%s", data_path, .)))) %>%
  unnest(data) %>%
  select(ID, count, one_of(old.names)) %>% # select columns from codebook in loaded data  
  setNames(c("ID", "count", new.names)) # rename columns with new names  
df1
```


## Descriptives: Within-Person Correlations  
With these kinds of data, the first thing, we may want to do is look at within-person correlations, which we can do with `purrr`.  

```{r}
nested.r <- df1 %>%
  group_by(ID) %>%
  nest() %>%
  mutate(r = map(data, ~cor((.) %>% select(-count), use = "pairwise")))
nested.r
```

We can access it like a list: 

```{r}
nested.r$data[[1]]
```

## Models  

To run separate models for each trait, we'll need to reshape our data.  

```{r}
df.long <- df1 %>%
  gather(key = item, value = value, -count, -ID, -satisfaction, na.rm = T)
df.long
```

To create composites, we'll separate traits from items.  
```{r}
df.long <- df1 %>%
  gather(key = item, value = value, -count, -ID, -satisfaction, na.rm = T) %>%
  separate(item, c("Trait",  "item"), sep = "_")
df.long
```


To create composites, we'll then `group_by()` trait, count, and ID and calculate the composites using `summarize()`  

```{r}
df.long <- df1 %>%
  gather(key = item, value = value, -count, -ID, -satisfaction, na.rm = T) %>%
  separate(item, c("Trait",  "item"), sep = "_") %>%
  group_by(ID, count, Trait) %>%
  summarize(value = mean(value)) 
df.long
```

Then we'll get within-person centered values using `scale()`.  
```{r}
df.long <- df1 %>%
  gather(key = item, value = value, -count, -ID, -satisfaction, na.rm = T) %>%
  separate(item, c("Trait",  "item"), sep = "_") %>%
  group_by(ID, count, Trait, satisfaction) %>%
  summarize(value = mean(value)) %>%
  group_by(ID, Trait) %>%
  mutate(value_c = as.numeric(scale(value, center = T, scale = F)))
df.long
```

And grand-mean centered within-person averages  
```{r}
df.long <- df.long %>%
  group_by(ID, Trait) %>%
  summarize(value_gmc = mean(value)) %>%
  group_by(Trait) %>%
  mutate(value_gmc = as.numeric(scale(value_gmc, center = T, scale = F))) %>%
  full_join(df.long)
df.long
```

### Run Models  
And now we are ready to run our models. But first, we'll `nest()` our data.  
```{r}
nested.mods <- df.long %>%
  group_by(Trait) %>%
  nest()
nested.mods
```

And now run the models.  
```{r}
nested.mods <- df.long %>%
  group_by(Trait) %>%
  nest() %>%
  mutate(model = map(data, ~lmer(satisfaction ~ value_c * value_gmc + (1 | ID), data = .)))
nested.mods
```


And get data frames of the results:  
```{r}
nested.mods <- df.long %>%
  group_by(Trait) %>%
  nest() %>%
  mutate(model = map(data, ~lmer(satisfaction ~ value_c * value_gmc + (1 | ID), data = .)),
         tidy = map(model, ~tidy(., conf.int = T)))
nested.mods
```

Which we can print into pretty data frames  
```{r}
nested.mods %>%
  select(Trait, tidy) %>%
  unnest(tidy)
```

### Plotting  
Which we can pretty easily turn into plots: 
```{r, fig.width= 3, fig.align='center'}
nested.mods %>%
  select(Trait, tidy) %>%
  unnest(tidy) %>%
  filter(term == "value_c") %>%
  ggplot(aes(x = Trait, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_errorbar(position = "dodge", width = .1) +
    geom_point(aes(color = Trait), size = 2) +
    geom_hline(aes(yintercept = 0), linetype = "dashed") +
    labs(y = "Personality State-\nSatisfaction Association") +
    coord_flip() +
    theme_classic() +
    theme(legend.position = "none")
```
