---
title: "Plotting Tutorial"
subtitle: "Applied Longitudinal Data Analysis"
author: "Emorie D Beck"
date: "`r Sys.setlocale('LC_TIME', 'C'); format(Sys.time(), '%d\\\\. %B %Y')`"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    code_folding: show
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, error = F)
```

<a href="https://raw.githubusercontent.com/emoriebeck/R-tutorials/master/ALDA/week_4_plotting/week_4_plotting.Rmd" download>Download .Rmd (won't work in Safari or IE)</a>  
<a href="https://github.com/emoriebeck/R-tutorials/tree/master/ALDA/week_4_plotting" target="_blank">See GitHub Repository</a>  

For a much more exhaustive tutorial, see <a href="https://raw.githubusercontent.com/emoriebeck/R-tutorials/master/mlm/Conditional_Models_doc.Rmd" download>this</a>.  

# Workspace  
## Packages  

First let's load in our packages. We're going to load in a couple of extras (`brms` and `tidybayes`) that we haven't used before. These will let us make some pretty plots later. 
```{r packages}
library(psych)
library(knitr)
library(kableExtra)
library(lme4)
library(broom.mixed)
library(brms)
library(tidybayes)
library(plyr)
library(tidyverse)

data_path <- "https://github.com/emoriebeck/R-tutorials/raw/master"
```

## Read in Data  
The National Longitudinal Study of Youths 1979 Child and Young Adult Sample (NLSYCYA) is a longitudinal study conducted by the National Bureau of Labor Statistics. The sample includes the children of the original 1979 sample, of which we will use a small subset. Here, we are going to use a subset of the more than 11,000 variables available that include the following:

Item Name   | Description                   | Time-Varying?
----------- | ----------------------------- | -------------
PROC_CID    | Participant ID                | No  
Dem_DOB     | Year of Date of Birth         | No
groups      | Jail, Community Service, None | No
DemPWeight  | Weight Percentile at age 10   | No
age         | Age of participant            | Yes
Year        | Year of Survey                | Yes
age0        | Age of participant (centered) | Yes
SensSeek    | Sensation-Seeking Composite   | Yes
CESD        | CESD Depression Composite     | Yes

```{r read data ex6}
load(url(sprintf("%s/ALDA/week_4_plotting/data/sample.RData", data_path)))

sample_dat
```


## Restructure Data  
These data are already largely cleaned as that is not our focus today. We just need to do a little bit of restructuring.  

To run our models using `purrr`, we need to restrucutre the data to long. While we're at it, we need to create a time variable centered at zero, so we can interpret our moderators later.  

```{r}
(df1_long <- sample_dat %>%
  gather(key = trait, value = value, CESD, SensSeek, na.rm = T) %>%
   mutate(wave = year - 1996,
          age0 = age-16)) 
```


# The Basic Growth Model  

We'll start with the basic growth model (just looking at change in a single variable over time).  

we will use list columns to do it. We'll start by using the `group_by()` and `nest()` functions from `dplyr` and `tidyr` to put the data for each trait into a cell of our data frame:  

```{r nest ex6}
(df1_nested <- df1_long %>%
  group_by(trait) %>%
  nest())
```

Now, our data frame is 2 x 2, with the elements in the second column each containing the data frame that corresponds to that trait. This makes it really easy to run our models using the `map()` family of unctions from `purrr`.  

Before we fit the full growth model, we will first fit the unconditional model. Below, we will add a new column to our data frame that will contain the unconditional model for each trait. 

In this case the model will be in the form of: 
<ul>
  <li> <strong>Level 1:</strong> $Y_{ij} = \beta_{0j} + \varepsilon{ij}$ </li>
  <li> <strong>Level 2:</strong>  </li>
    <ul> 
      <li> $\beta_{0j} = \gamma_{00} + U_{0j}$ </li>
    </ul>
</ul>

```{r fit0 ex6}
(df1_nested <- df1_nested %>%
  mutate(fit0 = map(data, ~lmer(value ~ 1 + (1 | PROC_CID), data = .))))
```

Now we can see we have a new list column in our data frame called fit0 that contains an S4 class lmerMod, which simply means your growth model. To understand model, I personally find it easiest to visualize it. What this model is telling us is the mean across all observations as well as the between-person variability in that estimate. 

Now, moving on to the growth model: 

In this case the model will be in the form of: 
<ul>
  <li> <strong>Level 1:</strong> $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{ij} + \varepsilon{ij}$ </li>
  <li> <strong>Level 2:</strong></li> 
    <ul>
      <li> $\beta_{0j} = \gamma_{00} + U_{0j}$ </li>
      <li> $\beta_{1j} = \gamma_{10} + U_{1j}$ </li>
    </ul>
</ul>

```{r fit12 ex6}
(df1_nested <- df1_nested %>%
  mutate(fit2 = map(data, ~lmer(value ~ 1 + age0 + (age0 | PROC_CID), data = .))))
```

Now that we've run our model, we are ready to plot the results.  

To do so, we'll write a short function that will get the predicted values both for group level effects, as well as for individual-level estimates. There's also a hidden function for getting standard errors of our pointwise estimates that we'll use to create confidence bands in our plots. Don't worry about that.  

```{r, echo = F}
pv_fun <- function(frame, m){
  nvar <- nrow(vcov(m)) # # of variables
  varnames <- str_replace_all(colnames(vcov(m)), "[()]", "") # names for vcov mat
  vcov_mat <- matrix(vcov(summary(m))@x, nvar) # chnage vcov mat class
  colnames(vcov_mat) <- varnames; rownames(vcov_mat) <- varnames # rename vcov mat
  frame_mat <- frame %>% as.matrix
  pv <- diag(frame_mat %*% vcov_mat %*% t(frame_mat)) # compute variance
  SE <- sqrt(pv) 
  return(SE)
}
```

## Predicted Values  
Now we'll use the predict function to get predicted values and the function I created to get pointwise standard errors for the fixed effects.    
```{r}
# function to get fixed effects
fixed_pred_fun <- function(m){
  frame <- tibble(Intercept = 1, age0 = seq(0, 8, .1)) %>%
    mutate(fixed_pred = predict(m, newdata = ., re.form = NA),
           age = age0 + 16)
  frame$SE <- pv_fun(frame %>% select(Intercept, age0), m)
  frame %>% select(-Intercept)
}

# function to get random effects predictions  
ran_pred_fun <- function(m){
  crossing(age0 = seq(0, 8, 1),
         PROC_CID = m@frame$PROC_CID) %>%
    mutate(ran_pred = predict(m, newdata = .),
           age = age0 + 16)
}

# get fixed and random efects, and also combine them.
(df1_nested <- df1_nested %>%
  mutate(fixed_pred = map(fit2, fixed_pred_fun),
         ran_pred = map(fit2, ran_pred_fun),
         combined_pred = map2(fixed_pred, ran_pred, full_join)))
```

## Plot  
Now let's 
```{r}
df1_nested %>%
  select(trait, combined_pred) %>%
  unnest(combined_pred) %>%
  ggplot(aes(x = age)) +
    geom_line(aes(y = ran_pred, color = trait, group = PROC_CID), size = .25, alpha = .5) +
    geom_line(aes(y = fixed_pred), size = 2) +
    lims(y = c(0,4)) +
    labs(x = "Age", y = "Composite Rating",
         title = "Simple Growth Models") +
    facet_grid(~trait) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none",
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

And with confidence bands:   

```{r}
df1_nested %>% 
  mutate(df = map_dbl(fit2, df.residual)) %>%
  select(trait, fixed_pred, df) %>%
  unnest(fixed_pred) %>%
  ggplot(aes(x = age, y = fixed_pred, fill = trait)) +
    geom_ribbon(aes(ymin=fixed_pred-1.96*SE,ymax=fixed_pred+1.96*SE),alpha=0.2,fill="blue") +  
    geom_line(size = 2) +
    facet_grid(~trait) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none",
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```


```{r}
df1_nested %>% 
  mutate(df = map_dbl(fit2, df.residual)) %>%
  select(trait, fixed_pred, df) %>%
  unnest(fixed_pred) %>%
  ggplot(aes(x = age, y = fixed_pred, fill = trait)) +
    stat_dist_lineribbon(
      aes(dist = "student_t", arg1 = unique(df), arg2 = fixed_pred, arg3 = SE),
      alpha = 1/4
    ) +
    facet_grid(~trait) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none",
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```


# Two-Level Categorical Moderator  

Let's start with the basic syntax:  

<ul>
  <li> <strong>Level 1:</strong> $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{1j} + \varepsilon{ij}$</li>
  <li> <strong>Level 2:</strong> </li>
    <ul> 
      <li> $\beta_{0j} = \gamma_{00} + \gamma_{01}*X_{2j} + U_{0j}$</li>
      <li> $\beta_{1j} = \gamma_{10} + \gamma_{11}*X_{2j} + U_{1j}$</li>
    </ul>
</ul>

Now let's swap that out for a 2 group sample from the present data:  

<ul>
  <li> <strong>Level 1:</strong> $Y_{ij} = \beta_{0j} + \beta_{1j}*age0_{ij} + \varepsilon{ij}$</li>
  <li> <strong>Level 2:</strong> </li>
    <ul> 
      <li> $\beta_{0j} = \gamma_{00} + \gamma_{01}*groupsNone + U_{0j}$</li>
      <li> $\beta_{1j} = \gamma_{10} + \gamma_{11}*groupsNone + U_{1j}$</li>
    </ul>
</ul>

Variable  | D1  
--------- | ---
Jail      | 0 
None      | 1

```{r}
df1_nested <- df1_nested %>%
  mutate(data_2g = map(data, function(x) x %>% filter(groups != "CommServ")),
         fit3 = map(data_2g, ~lmer(value ~ 1 + age0*groups + (age0 | PROC_CID), data = .)))
```

When we plot these, we are plotting the simple slopes. Subbing 1 and 0 into the equations above we end up with the following for the groups.  

None: $Y_{ij} = \gamma_{00} + \gamma_{10}*age$  
Jail: $Y_{ij} = \gamma_{00} + \gamma_{01} + (\gamma_{10} + \gamma_{11})*age$  

```{r}
fixed_pred_fun <- function(m){
  frame <- crossing(Intercept = 1, age0 = seq(0, 8, .1),
           groups = c("Jail", "None")) %>%
    mutate(fixed_pred = predict(m, newdata = ., re.form = NA),
           age = age0 + 16,
           groupsn = as.numeric(mapvalues(groups, unique(groups), c(1,0))),
           Int = age0*groupsn)
  frame$SE <- pv_fun(frame %>% select(Intercept, age0, groupsn, Int), m)
  frame %>% select(-Intercept, -groupsn, -Int)
}

fixed_pred_fun <- function(m){
  crossing(age0 = seq(0, 8, 1),
           groups = c("Jail", "None")) %>%
    mutate(fixed_pred = predict(m, newdata = ., re.form = NA),
           age = age0 + 16)
}

ran_pred_fun <- function(m){
  crossing(age0 = seq(0, 8, 1),
         PROC_CID = m@frame$PROC_CID) %>%
    left_join(m@frame %>% tbl_df %>% select(PROC_CID, groups)) %>%
    distinct() %>%
    mutate(ran_pred = predict(m, newdata = .),
           age = age0 + 16)
}

(df1_nested <- df1_nested %>%
  mutate(fixed_pred3 = map(fit3, fixed_pred_fun),
         ran_pred3 = map(fit3, ran_pred_fun),
         combined_pred3 = map2(fixed_pred3, ran_pred3, full_join)))
```

```{r}
df1_nested %>%
  select(trait, combined_pred3) %>%
  unnest(combined_pred3) %>%
  ggplot(aes(x = age)) +
    scale_color_manual(values = c("blue", "red")) +
    geom_line(aes(y = ran_pred, color = groups, group = PROC_CID), size = .25, alpha = .1) +
    geom_line(aes(y = fixed_pred, color = groups), size = 2) +
    lims(y = c(0,4)) +
    labs(x = "Age", y = "Composite Rating",
         title = "Simple Growth Models") +
    facet_grid(~trait) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none",
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

## Three-Level Categorical Moderator  
<ul>
  <li> <strong>Level 1:</strong> $Y_{ij} = \beta_{0j} + \beta_{1j}*age0_{ij} + \varepsilon{ij}$</li>
  <li> <strong>Level 2:</strong> </li>
    <ul> 
      <li> $\beta_{0j} = \gamma_{00} + \gamma_{01}*D1 + \gamma_{02}*D2 + U_{0j}$</li>
      <li> $\beta_{1j} = \gamma_{10} + \gamma_{11}*D1 + \gamma_{12}*D2 + U_{1j}$</li>
    </ul>
</ul>

Variable  | D1  | D2  
--------- | --- | ---
Jail      | 0   | 0
None      | 1   | 0
CommServ  | 0   | 1  

```{r}
df1_nested <- df1_nested %>%
  mutate(fit4 = map(data, ~lmer(value ~ 1 + age0*groups + (age0 | PROC_CID), data = .)))
```

When we plot these, we are plotting the simple slopes. Subbing 1 and 0 into the equations above we end up with the following for the groups.  

None: $Y_{ij} = \gamma_{00} + \gamma_{10}*age$  
Jail: $Y_{ij} = \gamma_{00} + \gamma_{01} + (\gamma_{10} + \gamma_{11})*age$  
Community Service: $Y_{ij} = \gamma_{00} + \gamma_{02} + (\gamma_{10} + \gamma_{12})*age$  

```{r}
fixed_pred_fun <- function(m){
  crossing(age0 = seq(0, 8, 1),
           groups = c("Jail", "None", "CommServ")) %>%
    mutate(fixed_pred = predict(m, newdata = ., re.form = NA),
           age = age0 + 16)
}

ran_pred_fun <- function(m){
  crossing(age0 = seq(0, 8, 1),
         PROC_CID = m@frame$PROC_CID) %>%
    left_join(m@frame %>% tbl_df %>% select(PROC_CID, groups)) %>%
    distinct() %>%
    mutate(ran_pred = predict(m, newdata = .),
           age = age0 + 16)
}

(df1_nested <- df1_nested %>%
  mutate(fixed_pred4 = map(fit4, fixed_pred_fun),
         ran_pred4 = map(fit4, ran_pred_fun),
         combined_pred4 = map2(fixed_pred4, ran_pred4, full_join)))
```

```{r}
df1_nested %>%
  select(trait, combined_pred4) %>%
  unnest(combined_pred4) %>%
  ggplot(aes(x = age)) +
    scale_color_manual(values = c("blue", "red", "black")) +
    geom_line(aes(y = ran_pred, color = groups, group = PROC_CID), size = .25, alpha = .1) +
    geom_line(aes(y = fixed_pred, color = groups), size = 2) +
    lims(y = c(0,4)) +
    labs(x = "Age", y = "Composite Rating",
         title = "Simple Growth Models") +
    facet_grid(~trait) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none",
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

## Continuous Time-Invariant Moderator  
```{r}
df1_nested <- df1_nested %>%
  mutate(fit5 = map(data, ~lmer(value ~ 1 + age0*DemPweight + (age0 | PROC_CID), data = .)))
```

```{r}
fixed_pred_fun <- function(m){
  desc <- Rmisc::summarySE(m@frame, measurevar = "DemPweight")
    crossing(age0 = seq(0, 8, 1),
           DemPweight = c(desc$DemPweight - desc$sd, 
                          desc$DemPweight, 
                          desc$DemPweight + desc$sd)) %>%
    mutate(fixed_pred = predict(m, newdata = ., re.form = NA),
           age = age0 + 16,
           Weight = factor(DemPweight, levels = unique(DemPweight), labels = c("-1SD", "0SD", "1SD")))
}

ran_pred_fun <- function(m){
  crossing(age0 = seq(0, 8, 1),
         PROC_CID = m@frame$PROC_CID) %>%
    left_join(m@frame %>% tbl_df %>% select(PROC_CID, DemPweight)) %>%
    distinct() %>%
    mutate(ran_pred = predict(m, newdata = .),
           age = age0 + 16)
}

(df1_nested <- df1_nested %>%
  mutate(fixed_pred5 = map(fit5, fixed_pred_fun),
         ran_pred5 = map(fit5, ran_pred_fun),
         combined_pred5 = map2(fixed_pred5, ran_pred5, full_join)))
```

```{r}
df1_nested %>%
  select(trait, combined_pred5) %>%
  unnest(combined_pred5) %>%
  ggplot(aes(x = age)) +
    scale_color_manual(values = c("blue", "red", "black")) +
    # geom_line(aes(y = ran_pred, group = PROC_CID), size = .25, alpha = .1) +
    geom_line(aes(y = fixed_pred, color = Weight), size = 1) +
    lims(y = c(0,4)) +
    labs(x = "Age", y = "Composite Rating",
         title = "Simple Growth Models") +
    facet_wrap(~trait, scales = "free_y") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none",
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

