---
title: "Conditional Models"
author: "Emorie D Beck"
date: "9/7/2017"
output: 
  pdf_document:
    keep_tex: yes
    toc: yes
    includes:
            in_header:
                header.tex
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    code_folding: show
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = T, cache = T, warning = F, message = F)
```

# Workspace 

## Packages
```{r}
library(psych)
library(broom)
library(plyr)
library(tidyverse)
```

## Data  
This week, our data are going to (continue to) come from the German Socioeconomic Panel Study (GSOEP). The GSOEP is a longitudinal study of adults in German housesholds. The study has a broad range of variables, but for our purposes we're just going to use personality ratings, life events, age, gender, and general life satisfaction from 2005 to 2015. We'll create a composite of life satisfaction for one of our tests of moderators. We'll use life events similarly to how we did for logistic regression, except this time as a moderator variable.    

We need to extend our codebook. To create it, go to https://data.soep.de/soep-core# and use the search feature to find the variables you need or https://data.soep.de/soep-long/topics/ where you can scroll through topics (this may be easier for finding the personality variables). Use your codebook from last week, and add the additional variables.   

Each year has several different files. Thankfully, for our purposes, we just need one file for each year. The first part of that file name indexes which wave it is. Waves are labeled a (1985) to bf (2015). Once the waves hit z, they start over at "ba". The second piece of the filename indexes which type of file it is. We need the "p" files, which stand for person. So, for instance, 2005 is "vp.sav".  

There are different ways to load it in, but I would recommend using some form of loop, which should do the following:  
1. read in the file for a specific year (e.g. using `haven::read_sav()`). 
2. pull the variables from the codebook from that year (e.g. using `select()`).
    - NOTE: you should pull certain variables, like the person and household IDs for every year.  
3. rename those variables in wide format.  
4. add a column to the data for that year that indexes what year the observation is.  
5. merge the data from that year with previous years.  

For help with this, see https://emoriebeck.github.io/R-tutorials/purrr/. I'll give you a `purrr` solution later in the week.    

Once you've got the codebook, we should be ready to go. 

```{r}
wd <- "https://github.com/emoriebeck/R-tutorials/blob/master/RA_Files/Week_7_Growth_Models_II"
# load your codebook 
destfile <- "Codebook_EDB.xlsx"
curl::curl_download(sprintf("%s/Codebook_EDB.xlsx?raw=true", wd), destfile)
codebook <- readxl::read_excel(destfile) %>%
  mutate(Item = stringr::str_to_lower(Item))
```

```{r}
all.old.cols <- (codebook %>% filter(class == "proc" & Year == 0))$Item
all.new.cols <- (codebook %>% filter(class == "proc" & Year == "0"))$new_name

# create short function to read in separate files for each wave
read_fun <- function(file, year){
  print(year)
  old.names <- (codebook %>% filter(Year == year & class %in% c("group", "predictor", "proc")))$Item
  new.names <- (codebook %>% filter(Year == year & class %in% c("group", "predictor", "proc")))$new_name
  z <- haven::read_sav(url(sprintf("%s/data/%sp.sav?raw=true", wd, file))) %>%
    select(one_of(all.old.cols), one_of(old.names)) %>%
    setNames(c(all.new.cols, new.names)) 
}

# you need letters, not numbers to index different data files. 
# but years will be more useful to index your codebook, so we'll 
# put both in our starting data frame. I've filled out this part. 
# Now you just need to figure out how use that to load the files 
# and get the correct variables (one's that repeat year to year)
dat <- tibble(
  Year = as.character(seq(2005, 2015,1)),
  file = c(letters[22:26], paste("b", letters[1:6], sep = ""))) %>%
  mutate(data = map2(file, Year, read_fun)) %>%
  unnest(data)
```

## Descriptives  
Because our data are now longitudinal, we need to split our descriptives by year. Try doing this using the `describeBy()` in the `psych` package.  
```{r}
# run the descriptives and check variable ranges
describeBy(dat, dat$Year)
```

## Check Missings 
How are missings coded in this data set? Do we need to make any changes to how they are coded?  
```{r}
# You should have noted some variables that needed "scrubbed" (changed to missing)
# change those to NA using your preferred method
dat <- dat %>% mutate_all(funs(mapvalues(., seq(-1,-7,-1), rep(NA,7), warn_missing = F)))
```

## Recode Variables  
```{r}
# You should have your keys. Reverse code the items that need reverse coded. 
keys     <- codebook$rev_code[codebook$rev_code == -1]
items    <- codebook$new_name[codebook$rev_code == -1]
dat[,items] <- reverse.code(keys, dat[,items], mini = 1, maxi = 7)

# I'm going to give you this chunk because apparently some people don't know what year they were born
dat <- dat %>% 
  group_by(PROC_SID) %>% 
  mutate(
    Dem_DOB = max(Dem_DOB, na.rm = T),
    Dem_DOB = ifelse(is.infinite(Dem_DOB) == T, NA, Dem_DOB),
    Dem_Sex = max(Dem_Sex, na.rm = T),
    Dem_Sex = ifelse(is.infinite(Dem_Sex) == T, NA, Dem_Sex)
  )
```

## Create New Variables  
For these data, we need to create an age variable. There isn't one in the data set.v
```{r}
# create an age variable by subtracting the date of birth from 2005 
# change gender to a factor 
dat <- dat %>% 
  mutate(age = 2005 - Dem_DOB,
         gender = factor(Dem_Sex, levels = c(1,2), labels = c("Male", "Female")))

# create a composite "parent died" variable
dat <- dat %>%
    group_by(PROC_SID) %>%
    mutate(LE_ParDied = max(LE_MomDied,LE_DadDied, na.rm = T),
           LE_ParDied = ifelse(is.nan(LE_ParDied) == T, NA, LE_ParDied)) 
```

## Create composites
For these data, we have lots of items, so we don't just want to create composites for the Big 5, we also want to create composites for the facets of each of the Big 5. Use the methods we learned before to do so.  

### Personality  
```{r}
pers_dat <- dat %>%
  gather(key = item, value = value, BF_A1:BF_O3, na.rm = T) %>%
  separate(item, c("trait", "item"), -1) %>%
  group_by(trait, PROC_SID, PROC_household, Year, Dem_DOB, age, Dem_Sex, gender) %>%
  summarize(value = mean(value, na.rm = T)) %>%
  spread(key = trait, value = value) %>% 
  ungroup()
```

### Life Events
We want to code if someone experienced a life event anywhere within the study period. Experiment with different ways of doing this. Note that we want to figure out if a participant EVER responded with a 1 to any of our life event variables. 
```{r}
le_dat <- dat %>% select(-contains("BF")) %>%
  gather(key = le, value = le_value, contains("LE_")) %>%
  group_by(PROC_SID, age, gender, le) %>%
  summarize(le_value = ifelse(any(le_value == 1), 1, 0),
            le_value = ifelse(is.na(le_value), 0, le_value)) %>%
  ungroup() %>%
  spread(key = le, value = le_value)
```

### Merge the files  
```{r}
dat_final <- pers_dat %>% full_join(le_dat)
```

# Adding Predictors in MLM/HLM  
Last week, we talked about building simple longitudinal models in R using MLM/HLM. This week, we're going to extend the "simple" MLM growth curve model to include additional predictors. Depending on the type of predictor, you can add these predictors at level 1 or level 2. Figuring out which level your covariates are at is very important. Part of this will have to do with whether you expect level 1 and level 2 variables to interact. 

To demonstrate what I mean by this, we can start by considering the case where we want to add a covariate like age at level 1, where we want to partial out how age influences the model, but not to see how personality changes over time as a function of age. 

The simple growth model has the following form: 
**Level 1**: $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{ij} + \varepsilon_{ij}$  
**Level 2**:  
Random Intercept: $\beta_{0j} = \gamma_{00} + U_{0j}$  
Random Slope: $\beta_{1j} = \gamma_{10} + U_{1j}$  

## Additional Level 1 Covariate  
When we add age as a covariate, it has the following form:  
**Level 1**: $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{ij} + \beta_{2j}*age_{ij} + \varepsilon_{ij}$  
**Level 2**:  
Intercept: $\beta_{0j} = \gamma_{00} + U_{0j}$ (random intercept)  
Time Slope: $\beta_{1j} = \gamma_{10} + U_{1j}$ (random slope)  
<font color = "blue">Age Slope: $\beta_{2j} = \gamma_{20}$ (no random slope)</font>  

Note that this does add an additional level 2 equation, but we don't add an addiitonal random slope. How does this affect our model? One of the easiest ways to see this is to "sub in" the level 2 models into the level 1 equation, which gives us:  
$Y_{ij} = \gamma_{00} + U_{0j} + (\gamma_{10} + U_{1j})*time_{ij} + (\gamma_{20})*age_{ij} + \varepsilon{ij}$  

So in this case, we are just adding one addition term (a fixed effect age term), which we would interpret the same as we would if we added an age variable in multiple regression. 

## Additional Level 2 Covariate  
Sometimes, we are interested in covariates that we expect to interact with our level 1 covariates, but we have no reason to think that we should 
**Level 1**: $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{ij} + \varepsilon_{ij}$  
**Level 2**:  
Intercept: $\beta_{0j} = \gamma_{00} + \gamma_{01}*life\_sat\_w1_{j} + U_{0j}$ (random intercept)  
Time Slope: $\beta_{1j} = \gamma_{10} + \gamma_{11}*life\_sat\_w1_{j} + U_{1j}$ (random slope)  

Which combines to:  
$Y_{ij} = \gamma_{00} + \gamma{11}*life\_sat\_w1_{j} + U_{0j} + (\gamma_{10} + \gamma{11}*life\_sat\_w1_{j} + U_{1j})*time_{ij} + \varepsilon{ij}$  
$Y_{ij} = \gamma_{00} + \gamma{11}*life\_sat\_w1_{j} + \gamma_{10}*time_{ij} + \gamma{11}*life\_sat\_w1_{j}*time_{ij} + U_{0j} + U_{1j}*time + \varepsilon_{ij}$  


## Simple Growth Curve Model  

\begin{itemize}
  \item \textbf{Level 1:} $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{ij} + \varepsilon{ij}$  
  \item \textbf{Level 2:} 
    \begin{itemize} 
      \item $\beta_{0j} = \gamma_{00} + U_{0j}$
      \item $\beta_{1j} = \gamma_{10} + U_{1j}$
    \end{itemize}
\end{itemize}

```{r, echo = F}
sample_dat %>%
  ggplot(aes(x = age, y = SensSeek)) + 
    geom_smooth(aes(group = PROC_CID), method = "lm", se = F, color = "gray", size = .2) +
    geom_smooth(method = "lm", se = F, color = "red", size = 2) +
    labs(x = "Age", y = "Sensation Seeking Composite", title = "Simple Growth Curve") +
    theme_classic() +
    theme(legend.position = "none",
          axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

## In R
\small
```{r}
mod0 <- lmer(SensSeek ~ age0 + (1|PROC_CID), data = sample_dat)
```

\centering
\small
```{r, echo = F}
summary(mod0)
```

\normalsize
\raggedright

## Conditional Models: Adding Predictors
Let's see if we can better predict participants' change in sensation seeking over time by adding covariates.  

Predictor      | Continuous  | Categorical 
----------- | ----------- | -----------
Time Invariant  | Weight for Age  | Group 
Time Varying  | CESD Scores | Depression  


# Time Invariant Predictors
## Time Invariant Predictors: Continuous

The basic equation, specifying a random intercept and slope:    
\begin{itemize}
  \item \textbf{Level 1:} $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{1j} + \varepsilon{ij}$
  \item \textbf{Level 2:} 
    \begin{itemize} 
      \item $\beta_{0j} = \gamma_{00} + \gamma_{01}*X_{2j} + U_{0j}$
      \item $\beta_{1j} = \gamma_{10} + \gamma_{11}*X_{2j} + U_{1j}$
    \end{itemize}
\end{itemize}

But we need to break this down to see that adding additional predictors results in interaction terms:  

$Y_{ij} = \gamma_{00} + \gamma_{01}*X_{2j} + U_{0j} + (\gamma_{10} + \gamma_{11}*X_{2j} + U_{1j})*X_{1j} + \varepsilon{ij}$
$Y_{ij} = \gamma_{00} + \gamma_{01}*X_{2j} + \gamma_{10}*X_{1j} + \textcolor{red}{\gamma_{11}*X_{2j}*X_{1j}} + U_{0j} + U_{1j}*X_{1j} + \varepsilon{ij}$

We can also fit this with intercepts depending on weight, but without the change (slope) dependent on weight:  
$Y_{ij} = \gamma_{00} + \gamma_{01}*X_{2j} + U_{0j} + (\gamma_{10} + U_{1j})*X_{1j} + \varepsilon{ij}$
$Y_{ij} = \gamma_{00} + \gamma_{01}*X_{2j} + \gamma_{10}*X_{1j} + U_{0j} + U_{1j}*X_{1j} + \varepsilon{ij}$

### Continuous Example - Weight for Age Percentile  

\small
```{r}
describe(sample_dat$DemPweight)
```

\normalsize

```{r, echo = F}
sample_dat %>%
  ggplot(aes(x = age, y = SensSeek)) + 
    geom_smooth(aes(group = PROC_CID, color = DemPweight), method = "lm", se = F, size = .2) +
    geom_smooth(method = "lm", se = F, color = "red", size = 1) +
    labs(x = "Age", y = "Sensation Seeking Composite", title = "Simple Growth Curve") +
    theme_classic() +
    theme(#legend.position = "none",
          axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

\small
```{r,results='hide'}
# time invariant covariate with random intecept (with weight as covariate) 
# and slope (without weight as a covariate)
mod1a <- lmer(SensSeek ~ age0 + DemPweight + (age0|PROC_CID), 
              data = sample_dat)

summary(mod1a)

# time invariant predictor with random slope and intercept
mod1b <- lmer(SensSeek ~ age0 + DemPweight + age0*DemPweight + 
                (age0|PROC_CID), data = sample_dat)

summary(mod1b)
```

\normalsize

##Time Invariant Predictors: Categorical 
### Categorical Example - 2 level group

Let's start with the basic syntax:  

\begin{itemize}
  \item \textbf{Level 1:} $Y_{ij} = \beta_{0j} + \beta_{1j}*time_{1j} + \varepsilon{ij}$
  \item \textbf{Level 2:} 
    \begin{itemize} 
      \item $\beta_{0j} = \gamma_{00} + \gamma_{01}*X_{2j} + U_{0j}$
      \item $\beta_{1j} = \gamma_{10} + \gamma_{11}*X_{2j} + U_{1j}$
    \end{itemize}
\end{itemize}

Now let's swap that out for a 2 group sample from the present data:  

\begin{itemize}
  \item \textbf{Level 1:} $Y_{ij} = \beta_{0j} + \beta_{1j}*age0_{ij} + \varepsilon{ij}$
  \item \textbf{Level 2:} 
    \begin{itemize} 
      \item $\beta_{0j} = \gamma_{00} + \gamma_{01}*groupsNone + U_{0j}$
      \item $\beta_{1j} = \gamma_{10} + \gamma_{11}*groupsNone + U_{1j}$
    \end{itemize}
\end{itemize}

Variable  | D1  
--------- | ---
Jail      | 0 
None      | 1

And plot it.  
```{r, echo = F}
sample_dat %>% filter(groups != "CommServ") %>%
  ggplot(aes(x = age, y = SensSeek, color = groups)) +
    geom_smooth(aes(group = PROC_CID),method = "lm", se = F, color = "gray", size = .2) +
    geom_smooth(method = "lm", se = F, size = 1) +
    labs(x = "Age", y = "Sensation Seeking Composite",
         title = "2 Group Time Invariant Conditional Growth Models") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```


And model it:  
\small
```{r}
mod2g <- lmer(SensSeek ~ age0 + DemPweight + groups + age0*groups + (age0|PROC_CID), 
              data = sample_dat %>% filter(groups != "CommServ"))
summary(mod2g)

source("https://raw.githubusercontent.com/emoriebeck/homeworks/master/table_fun.R")
options(knitr.kable.NA = '')
table_fun(mod2g)
```

\normalsize

### Categorical Example - 3 level group  

```{r, echo = F}
sample_dat %>%
  ggplot(aes(x = age, y = SensSeek, color = groups)) +
    geom_smooth(aes(group = PROC_CID), method = "lm", se = F, color = "gray", size = .2) +
    geom_smooth(method = "lm", se = F, size = 1) +
    labs(x = "Age", y = "Sensation Seeking Composite",
         title = "2 Group Time Invariant Conditional Growth Models") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```


\begin{itemize}
  \item \textbf{Level 1:} $Y_{ij} = \beta_{0j} + \beta_{1j}*age0_{ij} + \varepsilon{ij}$
  \item \textbf{Level 2:} 
    \begin{itemize} 
      \item $\beta_{0j} = \gamma_{00} + \gamma_{01}*D1 + \gamma_{02}*D2 + U_{0j}$
      \item $\beta_{1j} = \gamma_{10} + \gamma_{11}*D1 + \gamma_{12}*D2 + U_{1j}$
    \end{itemize}
\end{itemize}

Variable  | D1  | D2  
--------- | --- | ---
Jail      | 0   | 0
None      | 1   | 0
CommServ  | 0   | 1  


\small
```{r}
mod3g <- lmer(SensSeek ~ age0 + groups + age0*groups + 
                (age0|PROC_CID), data = sample_dat)
summary(mod3g)
```


\normalsize

# Side Notes: Practical Applications
## Side Note: `lme4` helper functions  

```{r, eval = F}
vcov(mod2g)
VarCorr(mod2g)
fixef(mod2g)
head(ranef(mod2g)[[1]])
head(coef(mod2g)[[1]])
confint.merMod(mod2g, method = "boot")
reghelper::ICC(mod2g)
MuMIn::r.squaredGLMM(mod2g)
```

\small
```{r}
vcov(mod2g)
```


\small
```{r}
VarCorr(mod2g)
```


\small
```{r}
fixef(mod2g)
```


\small
```{r}
head(ranef(mod2g)[[1]])
```

```{r}
head(coef(mod2g)[[1]])
```

\small
```{r, warning = F, message = F}
confint.merMod(mod2g, method = "boot", nsim = 10)
```

All units of the random effects are in standard deviation units (which means you need to square them to get the variance!!)  
\begin{itemize}
  \item .sig01 = sd of random intercept = $\sqrt{\tau_{00}}$  
  \item .sig02 = correlation between slope and intercept = $\sqrt{\tau_{10}}$  
  \item .sig03 = sd of random slope = $\sqrt{\tau_{11}}$  
  \item .sigma = residual variance = $\hat{\sigma}$  
\end{itemize}


\small
```{r}
reghelper::ICC(mod2g)
```

***
\small
\textbf{Conditional $R^2$:} How much variance fixed + random effects explain  
\textbf{Marginal $R^2$:} how much variance the fixed effects explain  

[explained here](https://jonlefcheck.net/2013/03/13/r2-for-linear-mixed-effects-models/)  

```{r}
MuMIn::r.squaredGLMM(mod2g)
```

\normalsize

## Side Note: Creating MLM Tables
There are lots of helpful packages for this, including `stargazer` and `sjPlot`, which are demonstrated below.  
\small
```{r, eval = F}
stargazer::stargazer(mod2g)
sjPlot::sjt.lmer(mod2g)
```

\normalsize

The problem is that `stargazer()` doesn't include all the terms we want, and `sjt.lmer()` only renders html. Embedded in the `.Rmd` version of these slides is some code that should help you to extract the terms you need and create a table using `dplyr` and `tidyr` that you can render in \LaTeX using `stargazer`.

But let's understand where those variables came from. To do so, we'll use the `broom` package in R to grab the terms we need.  

Description                 | Math Notation 
--------------------------- | ------------------
Fixed Effect Intercept      | $\gamma_{00}$  
Fixed Effect Group Intercept  | $\gamma_{01}$  
Fixed Effect Age Slope      | $\gamma_{10}$  
Fixed Effect Group Slope    | $\gamma_{11}$  
Individual Random Intercepts  | $U_{0j}$  
Variance of Random Intercepts | $\tau_{00}$  
Random Age Slopes | $U_{10}$  
Variance of Random Age Slopes | $\tau_{11}$  
Correlation b/w Random Slopes and Intercepts  | $\tau_{10}$  
Residual Variance | $\hat{\sigma}^2$   
Intraclass Correlation  | ICC   
Conditional $R^2$           | $R^2_c$  
Marginal $R^2$              | $R^2_m$  


```{r, results='hide'}
broom::tidy(mod2g)
broom::glance(mod2g)
```

\small
```{r, echo = F}
broom::tidy(mod2g)
broom::glance(mod2g)
```

Below is code that *should* work for all models. Just run the function and save it as an `R` object. You can use this with `papaja` and the `apa_table()` function pretty easily. The trick is that if you are not using the papaja template, the proper LaTeX packages may not be loaded. You can get around this by attaching a .tex file calling the packages under "in_header: header.tex" in your YAML header. The YAML header of this .Rmd file contains the necessary syntax and the header.tex file with the proper packages.  
```{r, echo = T, results = 'hide'}
## here's some code to make a table. You shouldn't need to modify anything here 
## I've also included an R file called table_fun.R that you can directly source from github.
## fixed effects first ##
table_fun <- function(model){
  fixed <- broom::tidy(model) %>% filter(group == "fixed") %>%
    select(term, estimate) 
  ## add random effects ##
  rand <- VarCorr(model)[[1]]
  if(nrow(rand) > 1){
  rand <- rand[1:nrow(rand), 1:nrow(rand)]
  }
  colnames(rand)[colnames(rand) == "(Intercept)"] <- "Intercept"
  rownames(rand)[rownames(rand) == "(Intercept)"] <- "Intercept"
  vars <- rownames(rand)
  rand[upper.tri(rand)] <- NA
  rand <- data.frame(rand) %>% mutate(var1 = rownames(.)) %>%
    gather(key = var2, value = estimate, -var1, na.rm = T) %>%
    mutate(var1 = mapvalues(var1, vars, 0:(length(vars)-1)),
           var2 = mapvalues(var2, unique(var2), 0:(length(vars)-1))) %>%
    filter(var1 == var2) %>%
    unite(var, var1, var2, sep = "") %>%
    mutate(var = sprintf("$\\tau_{%s}$", var))
  ## get confidence intervals ##
  CI <- data.frame(confint.merMod(model, method = "boot", nsim = 10, oldNames = F)) %>%
    mutate(term = rownames(.)) %>% setNames(c("lower", "upper", "term"))
  
  CI %>% filter(term == "sigma") %>%
    mutate(estimate = sigma(model),
           term = "$\\sigma^2$",
           type = "Residuals")
  
  ## Get ICC & R2 values ##
  ICC <- reghelper::ICC(model)
  R2 <- MuMIn::r.squaredGLMM(model)
  
  ## format the fixed effects
  fixed <- fixed %>% left_join(CI %>% filter(!grepl(".sig", term))) %>%
    mutate(type = "Fixed Parts")
  
  rand <- rand %>%
    left_join(
      CI %>% filter(grepl("sd", term)) %>%
        mutate(lower = lower^2, upper = upper^2,
               var = mapvalues(term, unique(term), 0:(length(unique(term))-1)),
               var = sprintf("$\\tau_{%s%s}$", var, var)) %>% select(-term)) %>%
    mutate(type = "Random Parts") %>% rename(term = var)
  
  mod_terms <- tribble(
    ~term, ~estimate, ~type,
    # "ICC", ICC, "Model Terms",
    "$R^2_m$", R2[1], "Model Terms",
    "$R^2_c$", R2[2], "Model Terms"
  )
  
  tab <- fixed %>%
    full_join(rand) %>%
    mutate(CI = sprintf("[%.2f, %.2f]", lower, upper)) %>%
    select(-lower, -upper) %>%
    full_join(mod_terms) %>%
    mutate(estimate = sprintf("%.2f", estimate)) %>%
    dplyr::rename(b = estimate) %>%
    select(type, everything())
  return(tab)
}
# you can use this with papaja and the apa_table function pretty easily
# the trick is that if you are not using the papaja template, the proper
# LaTeX packages may not be loaded. You can get around this by attaching
# a .tex file calling the packages under "in_header: header.tex" in your YAML
# header the YAML header of this .Rmd file contains the necessary syntax and 
# the header.tex file with the proper packages

tab <- table_fun(mod2g)

```

###Basic: `kable()`  
\small
```{r}
options(knitr.kable.NA = '')
knitr::kable(tab, caption = "Ugly MLM Table Example")
```


### More Advanced: `kable()` + `kableExtra`  
\small
```{r, results = 'asis'}
library(kableExtra)
options(knitr.kable.NA = '')
tab %>% #select(-type) %>%
    mutate(term = gsub("[()]", "", term)) %>% 
  knitr::kable(.,
    caption = "Not Quite Right kableExtra MLM Table Example", 
    format = "html", 
    booktabs = T, 
    escape = F) %>%
  group_rows("Fixed", 1,4) %>%
  group_rows("Random", 5,9) %>%
  group_rows("Model", 9,11) %>%
  collapse_rows(1) %>%
  kable_styling(full_width = F)
  add_header_above(c(" ", " ", "Model 1" = 2))
```


### Alternative: `papaja` + `apa_table()`  
\small
```{r, results = 'asis'}
papaja::apa_table(tab %>% select(-type),caption = "papaja MLM Table Example", 
    na_string = "", stub_indents = list(Fixed = c(1:4), Random = c(5:8), Summary = c(9:11)),
    col_spanners = list(`Sensation Seeking` = c(2,3)))
```
\normalsize

###The Case of Multiple Models
Basically you can run this multiple times and then join the data frames together. The only trick is that you'll have to change the column names to make sure they are distinct. My favorite way of doing this is with the `purrr` package in R (anyone who knows me can feign shock at this). So basically, you make a dataframe of your models, which looks like this:
```{r, results = 'asis'}
mod.df <- tibble(
  outcome = c("Model 2g", "Model 3g", "Model Cont"), # just plug in whatever name for your outcome
  mod = c(mod2g, mod3g, mod1b) # basically just listing models that correspond to your outcomes
)

# then you use purrr

mod.df.long <- mod.df %>%
  mutate(tab = map(mod, table_fun)) %>%
  unnest(tab, .drop = T) 

table.df <- mod.df.long %>%
  gather(key = param, value = value, b, CI) %>%
  unite(param, outcome, param, sep = ".") %>%
  spread(key = param, value = value) %>%
  mutate(type = factor(type, levels = c("Fixed Parts", "Random Parts", "Model Terms"))) %>%
  arrange(type)

# then you can use apa_table or kable + kableExtra to make them pretty
options(papaja.na_string = " ")
papaja::apa_table(table.df %>% select(-type), caption = "Multi-Model Table",
          col_spanners = list(`2 group` = c(2,3), `3 group` = c(4,5), `Continuous` = c(6,7)),
          stub_indents = list(Fixed = seq(1,8), Random = seq(9,12), `Model Terms` = seq(13,15)),
          na_string = " ", 
          col.names = c(" ", rep(c("b", "CI"), times = 3)),
          align = c("l", rep("c", 6)))
```


##Side Note: Plotting
### Lazy Method: `sjPlot` + `sjt.int()`  

####Categorical
\small
```{r}
sjp.int(mod2g, type = "eff", p.kr = F, swap.pred = T)
```

####Continuous
\small
```{r}
sjp.int(mod1b, type = "eff", p.kr = F, swap.pred = T, mdrt.values = "meansd")
```

###Medium Advanced: Using the `predict()` function
####Categorical
```{r}
fixed.frame <- 
  data.frame(expand.grid(age0 = seq(0,10,2), 
             groups = c("None","Jail"))) %>%
  mutate(pred = predict(mod2g, newdata = ., re.form = NA))

fixed.frame %>%
  mutate(age = age0 + 14) %>%
  ggplot(aes(x = age, y = pred, color = groups)) +
    geom_line(size = 2) + 
    lims(y = c(0,4)) +
    labs(x = "Age", y = "Sensation Seeking Composite",
         title = "2 Group Time Invariant Conditional Growth Models") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))

```


####Continuous
```{r}
fixed.frame <- sample_dat %>%
  summarise(mean = mean(DemPweight, na.rm = T), 
            sd = sd(DemPweight, na.rm = T))

fixed.frame <- 
  data.frame(
    expand.grid(
      # here, you add values for your time variable and predictors
      age0 = seq(0,10,2), 
      DemPweight = c(fixed.frame$mean-fixed.frame$sd,
                     fixed.frame$mean,
                     fixed.frame$mean+fixed.frame$sd))) %>%
  mutate(pred = predict(mod1b, newdata = ., re.form = NA))


fixed.frame %>%
  mutate(Weight = factor(DemPweight, levels = unique(DemPweight), labels = c("-1SD", "0SD", "1SD")),
         age = age0 + 14) %>%
  ggplot(aes(x = age, y = pred, color = Weight)) +
    geom_line(size = 2) + 
    labs(x = "Age", y = "Sensation Seeking Composite",
         title = "Continuous Invariant Conditional Growth Models") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```



### More advanced: `expand.grid()` + `fixef()` + `ggplot2()`  
#### Categorical
```{r, echo=T, eval=T}
# example for categorical
fixed.frame <- 
    crossing(
      # here, you add values for your time variable and predictors
      Intercept = 1,
      age0 = seq(0,10,2), 
      groupsNone = c(0,1)) %>%
  # now take care of interactions and add an intercept
  mutate(`age0:groupsNone` = age0*groupsNone)
```

Before we move on, we need to calculate a few extra values. Why? Well, I like to include confidence bands around around my prediction lines, so we're going to do some weird stuff to extract what we need to do that.  
```{r}
# multiplying to get values for model frame
nvar <- nrow(vcov(mod2g)) # # of variables
varnames <- str_replace_all(colnames(vcov(mod2g)), "[()]", "") # names for vcov mat
vcov2g <- matrix(vcov(summary(mod2g))@x, nvar) # chnage vcov mat class
colnames(vcov2g) <- varnames; rownames(vcov2g) <- varnames # rename vcov mat
vcov2g <- vcov2g[,-3]; vcov2g <- vcov2g[-3,]
mat2g <- as.matrix(data.frame(unclass(fixed.frame))) # change model frame to matrix

pv2g <- diag(mat2g %*% vcov2g %*% t(mat2g)) # compute variance
fixed.frame$SE <- sqrt(pv2g) 
fixed.frame$SE2 <- sqrt(pv2g+sigma(mod2g)^2)
```


Cool, we've got the standard errors we need to calculate the CI band for each point now. Moving on to predicted values, let's do a little bit of matrix algebra to get our predicted values.  
```{r}
fixed.frame$pred <- as.vector(as.matrix(fixed.frame) %*% fixef(mod2g)[-3])
# minus 3 removes covariate

fixed.frame %>%
  mutate(groups = factor(groupsNone, levels = c(0,1), labels = c("Jail", "None")),
         age = age0 + 14) %>%
  ggplot(aes(x = age, y = pred, group = groups)) +
  geom_ribbon(aes(ymin=pred-2*SE,ymax=pred+2*SE),alpha=0.2,fill="blue") +  
  geom_line(aes(color = groups), size = 2) + 
    labs(x = "Age", y = "Sensation Seeking Composite",
         title = "2 Group Time Invariant Conditional Growth Models") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

####Continuous
Let's do the same thing with a continuous predictor. This is a little more complicated because we again have to calculate the mean and standard deviation to create our "new data."  
```{r, echo = T, eval = T}
detach(package:plyr)    
library(dplyr)
# example for continuous
fixed.frame <- sample_dat %>%
  group_by(PROC_CID) %>%
  summarise(DemPweight = mean(DemPweight, na.rm = T)) %>%
  ungroup() %>%
  summarise(mean = mean(DemPweight, na.rm = T), 
            sd = sd(DemPweight, na.rm = T))

fixed.frame <- 
  data.frame(
    expand.grid(
      # here, you add values for your time variable and predictors
      age0 = seq(0,10,2), 
      DemPweight = c(fixed.frame$mean-fixed.frame$sd,
                     fixed.frame$mean,
                     fixed.frame$mean+fixed.frame$sd))) %>%
  # now take care of interactions and add an intercept
  mutate(`age0:DemPweight` = age0*DemPweight,
         Intercept = 1) %>%
  # reordering everything
  select(Intercept, everything())

# multiplying to get values for model frame
fixed.frame$value <- as.vector(as.matrix(fixed.frame) %*% fixef(mod1b))

fixed.frame %>%
  mutate(Weight = factor(DemPweight, levels = unique(DemPweight), labels = c("-1SD", "0SD", "1SD")),
         age = age0 + 14) %>%
  ggplot(aes(x = age, y = value, color = Weight)) +
    geom_line(size = 2) + 
    labs(x = "Age", y = "Sensation Seeking Composite",
         title = "Continuous Invariant Conditional Growth Models") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))

```


\normalsize  

### Random Effects
We often also want to plot random effects to look at individual level trajectories / variability. Again, we need to use the `predict()` function to get these values. You can also totally use matrix algebra if you're like me and want to know what's happening "under the hood."  
#### Categorical
```{r}
tbl_df(mod3g@frame) %>% select(PROC_CID, age0, groups) %>%
  mutate(pred = predict(mod3g, newdata = .)) %>%
  ggplot(aes(x = age0, y = pred, group = PROC_CID, color = groups)) +
    geom_line(size = .5, alpha = .5) +
    labs(x = "Age", y = "Predicted Sensation Seeking", 
         title = "Predicted Random Effect Trajectories\nfor the 3 Group Model") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

#### Categorical
```{r}
mod1b@frame %>% select(-SensSeek) %>%
  mutate(pred = predict(mod1b)) %>%
  ggplot(aes(x = age0, y = pred, group = PROC_CID, color = DemPweight)) +
    geom_line(size = .5, alpha = .5) +
    labs(x = "Age", y = "Predicted Sensation Seeking", 
         title = "Predicted Random Effect Trajectories\nfor the 3 Group Model") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```

