---
title: "Samples and Sampling Distributions"
author: "Emorie Beck"
date: "2/16/2018"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, cache = T)
```

```{r global, include=FALSE}
library(tidyverse)
library(stringr)
# load data in 'global' chunk so it can be shared by all users of the dashboard
data_path <- "https://raw.githubusercontent.com/emoriebeck/Psy300_sampling_distributions/master"
cols <- c("DOB", "Fave_Color", "Initials", "Section", "Age", "Height", 
           "Weight", "Sleep", "Grade", "Good_Math", "Nervous_Stats")
dat <- read.csv(url(sprintf("%s/demo_survey.csv", data_path)), stringsAsFactors = F) %>%
  tbl_df() %>%
  filter(!(row_number() %in% c(1,2))) %>%
  select(Q67:Q86) %>%
  setNames(cols) 


dat <- dat %>%
  mutate(Height = str_replace_all(Height, "’", "'"),
         Height = str_replace_all(Height, "”", ""),
         Height = str_replace_all(Height, " ", ""),
         Height = str_replace_all(Height, '"', ''),
         Height = str_replace_all(Height, "''", ""),
         Height = str_replace_all(Height, "'$", ""),
         Height = str_replace_all(Height, "-", "'")) %>%
  separate(Height, c("Height_ft", "Height_in"), sep = "'") %>%
  mutate_at(vars(Section:Nervous_Stats), funs(as.numeric)) %>%
  mutate(Height_in = ifelse(Height_ft > 12, Height_ft, Height_in),
         Height_ft = ifelse(Height_ft > 12, 0, Height_ft),
         Height = Height_ft*12 + Height_in)
  
```

```{r, results = 'asis'}
library(knitr)
library(kableExtra)

dat %>% select(Initials, Section, Age, Height) %>%
  kable(., "html", booktabs = T, digits = 2, escape = F) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  scroll_box(width = "750px", height = "300px")
```


And we can also plot a distribution for this *population*.  
```{r}
dat %>% 
  ggplot(aes(x = Height)) +
    geom_histogram(binwidth = 1, color = "black", fill = "gray") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)))
```

# Sample Distributions
We can also take samples from our population (the class) of height.

## Raw values
```{r}

sample1 <- dat %>% select(Initials, Section, Age, Height) %>%
  filter(row_number() %in% sample(1:nrow(dat),15))

sample1 %>%
  kable(., "html", booktabs = T, digits = 2, escape = F) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  scroll_box(width = "750px", height = "150px")


sample2 <- dat %>% select(Initials, Section, Age, Height) %>%
  filter(row_number() %in% sample(1:nrow(dat),15)) 

sample2 %>%
  kable(., "html", booktabs = T, digits = 2, escape = F) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  scroll_box(width = "750px", height = "150px")
```


## Distributions
And look at their distributions.

```{r}
sample1 %>% mutate(sample = "Sample 1") %>%
  full_join(sample2 %>% mutate(sample = "Sample 2")) %>%
  ggplot(aes(x = Height, fill = sample)) +
    geom_histogram(binwidth = 1, color = "black") +
    facet_grid(.~sample) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none")
```

## Summary Statistics
And summary statistics:
```{r}
sample1 %>% mutate(sample = "Sample 1") %>%
  full_join(sample2 %>% mutate(sample = "Sample 2")) %>%
  group_by(sample) %>%
  summarize(Mean = mean(Height, na.rm = T),
            SD = sd(Height, na.rm = T),
            Min = min(Height, na.rm = T),
            Max = max(Height, na.rm = T))  %>%
  kable(., "html", booktabs = T, digits = 2, escape = F) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) 
```

# 100 Samples
They're pretty close. What if we did this 100 times?

## Raw values
```{r}
sample_fun <- function(df, size){
  dat %>% select(Initials, Section, Age, Height) %>%
    filter(row_number() %in% sample(1:nrow(dat),size)) 
}

sampled <- tibble(Sample = 1:100) %>%
  mutate(data = map(Sample, ~sample_fun(., 15)))
sampled %>%
  unnest(data) %>%
  group_by(Sample) %>%
  summarize(Mean = mean(Height, na.rm = T),
            SD = sd(Height, na.rm = T),
            Min = min(Height, na.rm = T),
            Max = max(Height, na.rm = T))  %>%
  kable(., "html", booktabs = T, digits = 2, escape = F) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = T) %>%
  scroll_box(width = "750px", height = "300px")
```

## Distribution
The means vary some. What if we look at the distribution in a histogram?
```{r}
sampled %>% 
  unnest(data) %>%
  filter(Sample %in% sample(1:100, 9)) %>%
  ggplot(aes(x = Height, fill = factor(Sample))) +
    geom_histogram(binwidth = 1, color = "black") +
    facet_wrap(~Sample, nrow = 3) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none")
```

## Sampling Distribution
And now what if we look at a distribution of the sample means?
```{r}
sampled %>% 
  unnest(data) %>%
  group_by(Sample) %>%
  summarize(Mean = mean(Height, na.rm = T)) %>%
  ggplot(aes(x = Mean)) +
    geom_histogram(binwidth = 1, color = "black", fill = "gray") +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)))
```

That looks more normal than the distribution of values in each sample.

# Varying Sample Size
Let's play with the number of samples.

```{r}
sampling_fun <- function(size, sample_size){
  tibble(Sample = 1:size) %>%
    mutate(data = map(Sample, ~sample_fun(., sample_size))) %>%
    unnest(Sample)
}

sampled2 <- expand.grid(
  Num_Samples = c(10, 100, 1000, 10000),
  Sample_Size = 15) %>%
  tbl_df %>%
  mutate(samples = map2(Num_Samples, Sample_Size, sampling_fun))
```

## Distributions
```{r}
sampled2 %>% unnest(samples) %>% unnest(data) %>%
  group_by(Num_Samples, Sample) %>%
  summarize(Mean = mean(Height, na.rm = T)) %>%
  ggplot(aes(x = Mean, y = ..density.., fill = factor(Num_Samples))) +
    geom_histogram(color = "black", bins = 20) +
    facet_grid(.~Num_Samples) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none")
```

Notably, our originally non-normal data now looks normal!!

## Standard Error of the Mean
And we can get something important from this: standard errors, which are the standard deviation of the sampling distribution of the mean.

```{r}
sampled2 %>% unnest(samples) %>% unnest(data) %>%
  group_by(Num_Samples, Sample) %>%
  summarize(Mean = mean(Height, na.rm = T)) %>%
  group_by(Num_Samples) %>%
  summarize(SD = sd(Mean, na.rm = T))
```

Don't worry about the fact that the standard errors get larger. Note instead that once we hit 1000 and 10000, they are consistent -- they are better estimates.  

# Central Limit Theorum

But wait, aren't standard errors supposed to get smaller? Yes but with larger SAMPLE SIZES, not more samples.  Watch:

```{r}
sample_fun <- function(df, size){
  dat %>% select(Initials, Section, Age, Height) %>%
    filter(row_number() %in% sample(nrow(dat), size, replace = T)) 
}

sampled3 <- expand.grid(
  Num_Samples = c(10, 100, 1000, 10000),
  Sample_Size = c(10,100,250),
  stringsAsFactors = F) %>%
  tbl_df %>%
  mutate(data = map2(Num_Samples, Sample_Size, sampling_fun))

sampled3 %>%
  unnest(data) %>% unnest(data) %>%
  filter(Num_Samples == 100) %>%
  group_by(Sample, Sample_Size) %>%
  summarize(Mean = mean(Height, na.rm = T)) %>%
  ggplot(aes(x = Mean,  fill = factor(Sample_Size))) +
    geom_histogram(bins = 50) +
    facet_grid(.~Sample_Size) +
    theme_classic() +
    theme(axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          legend.position = "none")
```

And this is the central limit theorum. Our estimates more closely approximate reality and even non-normal distributions of real data approximate normal when looking at means. Plus, when we look at the SD's, they've gotten A LOT narrower.  

```{r}
sampled3 %>% unnest(data) %>% unnest(data) %>%
  filter(Num_Samples == 100) %>%
  group_by(Sample_Size, Sample) %>%
  summarize(Mean = mean(Height, na.rm = T)) %>%
  group_by(Sample_Size) %>%
  summarize(SD = sd(Mean, na.rm = T))
```



